---
title: "The mammaPrintData package: annotated gene expression data from the Glas and Buyse breast cancer cohorts"
author:
  - name: Luigi Marchionni
    affiliation: The Sidney Kimmel Comprehensive Cancer Center, Johns Hopkins University School of Medicine
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: mammaPrintData
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Working with the mammaPrintData package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
options(width = 90)
```

# Overview

This vignette documents the `r Biocpkg("mammaPrintData")` package, which
provides the gene expression data of the Glas and Buyse breast cancer
cohorts [@Glas2006; @Buyse2006] as annotated `r Biocpkg("limma")` `RGList`
objects. These data were used to translate the 70-gene prognostic signature
of van 't Veer and Van de Vijver and colleagues [@vantVeer2002; @Vijver2002]
into the MammaPrint assay [@Glas2006], and to validate it independently in
the TRANSBIG multicenter cohort [@Buyse2006]. While the original studies
used the two-color Agilent/Rosetta 24k array, the redeveloped assay is based
on the custom 1.9k MammaPrint array (ArrayExpress array design
`A-MEXP-318`). For a review of these breast cancer studies see Marchionni et
al. [@Marchionni2007b; @Marchionni2008a].

The package contains:

- `glasRGcy5` and `glasRGcy3`: the **Glas cohort** (ArrayExpress series
  `E-TABM-115`, 162 patients), which combined a large proportion of the
  cases of the original van 't Veer and Van de Vijver cohorts and was used
  to implement the signature;
- `buyseRGcy5` and `buyseRGcy3`: the **Buyse cohort** (ArrayExpress series
  `E-TABM-77`, 307 patients), the European multicenter validation set;
- `seventyGeneSignature`: the 231 prognostic reporters of Table S2 of van
  't Veer et al. [@vantVeer2002], with the 70-gene signature flagged.

In both studies a two-color dye-swap design was applied, comparing each
sample against the same reference RNA pool ("MRP"). Therefore two distinct
`RGList` objects are provided for each cohort, one for each set of dye-swap
hybridizations. The objects contain the **unprocessed** intensities as
obtained from the raw ArrayExpress files: the pre-processed and normalized
values used by the authors were not part of the ArrayExpress submissions,
hence normalization is required prior to any analysis.

## Where the raw data live

Up to version 1.49.1 the original ArrayExpress files (about 210 MB) were
bundled in the package sources. They are now archived on Zenodo
(doi:[ZENODO_DOI]) and remain available per file from EBI BioStudies, the
current home of ArrayExpress [@Brazma2003]. The function
`fetchMammaPrintRaw()` retrieves them from either source (see Section
\@ref(raw)). The complete code used to generate the processed objects from
the raw files is kept in the `data-raw/` directory of the package sources
(<https://github.com/marchionniLab/mammaPrintData/tree/devel/data-raw>) and
is summarized in Section \@ref(generation).

# The processed data sets

Data sets are lazy-loaded: after attaching the package they can be referred
to by name. `r Biocpkg("limma")` provides the methods (`dim()`, `print()`,
subsetting, ...) for `RGList` objects.

```{r load}
library(mammaPrintData)
library(limma)
```

## Glas cohort

```{r glas}
class(glasRGcy5)
dim(glasRGcy5)
dim(glasRGcy3)
names(glasRGcy5)
```

The SDRF table of `E-TABM-115` available from ArrayExpress has one row per
hybridization rather than one per channel: the rows describing the patient
RNA (labeled with Cy5) carry the clinical information, whereas the rows
describing the reference pool (labeled with Cy3) do not. Consequently,
`glasRGcy5$targets` holds the patient-level information, while
`glasRGcy3$targets` holds the reference-pool metadata; the arrays of the two
objects are paired by position.

```{r glasTargets}
head(glasRGcy5$targets[, c(
  "Sample.Name", "Scan.Name", "putativeCohort",
  "OS", "OSevent", "TTM", "TTMevent", "FiveYearMetastasis"
)])
head(glasRGcy3$targets[, c("Sample.Name", "Label", "Scan.Name", "putativeCohort")])
```

## Buyse cohort

```{r buyse}
class(buyseRGcy5)
dim(buyseRGcy5)
dim(buyseRGcy3)
```

For `E-TABM-77` the SDRF table is complete (one row per channel), so the
`targets` component of both objects combines the patient and the reference
rows of each hybridization; the `Cy5` and `Cy3` columns record which sample
was labeled with which dye. Arrays are named after the patient identifier
in both objects.

```{r buyseTargets}
head(buyseRGcy5$targets[, c(
  "Sample.Name", "Cy5", "Cy3", "Characteristics.BioSourceProvider",
  "Factor.Value.MammaPrint.prediction", "OS", "OSevent",
  "TTM", "TTMevent", "FiveYearRecurrence", "toExclude"
)])
identical(colnames(buyseRGcy5), colnames(buyseRGcy3))
```

## Feature annotation

The `genes` component is the same for all four objects: the `A-MEXP-318`
array design annotation, plus the mapping of each feature to the 231-gene
and 70-gene prognostic lists (`genes231`, `genes70`) and the original
van 't Veer correlation with outcome (`gns231Cors`), where available.

```{r genes}
head(glasRGcy5$genes)
colSums(glasRGcy5$genes[, c("genes231", "genes70")])
identical(glasRGcy5$genes, buyseRGcy5$genes)
```

## The 70-gene signature

`seventyGeneSignature` is the parsed Table S2 of van 't Veer et al.
[@vantVeer2002], ordered by decreasing absolute correlation; according to
the supplementary methods of that paper the optimal signature consists of
the top 70 reporters.

```{r signature}
head(seventyGeneSignature)
table(seventyGeneSignature$signature70)
```

# Retrieving the raw data {#raw}

`fetchMammaPrintRaw()` downloads the original files once and caches them
with `r Biocpkg("BiocFileCache")`. Zenodo is the recommended source: it
serves one archive per ArrayExpress accession with the SDRF, the array
design and all raw files exactly as originally retrieved.

```{r fetchZenodo, eval = FALSE}
## The Glas cohort archive (70 MB) and the Buyse cohort archive (134 MB)
glasArchive <- fetchMammaPrintRaw("E-TABM-115")
buyseArchive <- fetchMammaPrintRaw("E-TABM-77")
rawDir <- tempfile("mammaPrintRaw_")
untar(glasArchive, exdir = rawDir)
untar(buyseArchive, exdir = rawDir)
list.files(rawDir)
## The van 't Veer supplementary spreadsheet
xls <- fetchMammaPrintRaw("seventyGene")
```

BioStudies serves the files individually (uncompressed). By default only
the SDRF and the array design are fetched; raw files can be requested by
name, or all at once (several hundred files per accession).

```{r fetchBioStudies, eval = FALSE}
meta <- fetchMammaPrintRaw("E-TABM-115", source = "biostudies")
sdrf <- read.delim(meta[["E-TABM-115.sdrf.txt"]], check.names = FALSE)
head(sdrf[["Array Data File"]])
## a couple of raw files ...
fetchMammaPrintRaw("E-TABM-115",
  source = "biostudies",
  files = sdrf[["Array Data File"]][1:2]
)
## ... or the complete set
fetchMammaPrintRaw("E-TABM-115", source = "biostudies", files = "all")
```

# How the objects were generated {#generation}

The code below is not evaluated when the vignette is built; it is a
condensed version of the scripts in `data-raw/`, which reproduce the shipped
objects exactly from the Zenodo archives.

## The 70-gene signature

```{r genSignature, eval = FALSE}
library(readxl)
gns231 <- read_xls(fetchMammaPrintRaw("seventyGene"))
colnames(gns231) <- gsub("\\s#", "", colnames(gns231))
colnames(gns231) <- gsub("\\s", ".", colnames(gns231))
gns231 <- gns231[, -grep("sp_xref_keyword_list", colnames(gns231))]
gns231 <- gns231[order(abs(gns231$correlation), decreasing = TRUE), ]
gns231$signature70 <- seq_len(nrow(gns231)) <= 70
seventyGeneSignature <- as.data.frame(gns231)
```

## Sample information and dye-swap sets

The SDRF tables are reduced to their informative columns. For the Glas
cohort the two dye-swap sets are identified from the label of the row
describing each hybridization; for the Buyse cohort the patient and
reference rows of each hybridization are combined.

```{r genTargets, eval = FALSE}
readSDRF <- function(file) {
  targets <- read.table(file,
    comment.char = "", sep = "\t", header = TRUE,
    stringsAsFactors = FALSE
  )
  targets <- targets[, apply(targets, 2, function(x) length(unique(x)) != 1)]
  targets <- targets[order(targets$Array.Data.File, decreasing = TRUE), ]
  colnames(targets) <- gsub("\\.$", "", gsub("\\.\\.", ".", colnames(targets)))
  targets
}
## Glas: one row per hybridization
targets <- readSDRF(file.path(rawDir, "ETABM115", "E-TABM-115.sdrf.txt.gz"))
targetsCy3info <- targets[targets$Source.Name == "MRP" & targets$Label == "Cy3", ]
targetsCy5info <- targets[targets$Source.Name != "MRP" & targets$Label == "Cy5", ]
## Buyse: one row per channel, combined per hybridization (see data-raw/buyseRG.R)
```

## Reading the raw intensities

```{r genRG, eval = FALSE}
colsList <- list(
  Rf = "Feature Extraction Software:rMedianSignal",
  Rb = "Feature Extraction Software:rBGMedianSignal",
  Gf = "Feature Extraction Software:gMedianSignal",
  Gb = "Feature Extraction Software:gBGMedianSignal",
  logRatio = "Feature Extraction Software:LogRatio",
  logRatioError = "Feature Extraction Software:LogRatioError"
)
RGcy5 <- read.maimages(paste0(targetsCy5info$Array.Data.File, ".gz"),
  source = "generic", columns = colsList,
  annotation = "Reporter identifier",
  path = file.path(rawDir, "ETABM115")
)
RGcy5$targets <- targetsCy5info
RGcy5$genes$ID <- gsub(".+A-MEXP-318\\.", "", RGcy5$genes[, "Reporter identifier"])
```

## Feature annotation

The `A-MEXP-318` array design file is parsed and each feature is mapped to
the 231-gene and 70-gene lists by accession number or gene symbol.

```{r genGenes, eval = FALSE}
genes <- read.table(file.path(rawDir, "ETABM115", "A-MEXP-318.adf.txt.gz"),
  skip = 21, sep = "\t", header = TRUE, stringsAsFactors = FALSE
)
colnames(genes) <- gsub("\\.$", "", gsub("\\.\\.", ".", colnames(genes)))
ids231 <- unique(c(seventyGeneSignature$accession, seventyGeneSignature$gene.name))
ids70 <- with(seventyGeneSignature, unique(c(accession[signature70], gene.name[signature70])))
genes$genes231 <- genes$Comment.AEReporterName %in% ids231
genes$genes70 <- genes$Comment.AEReporterName %in% ids70
## ... add the original correlations, then attach to the RGList
## (features are reordered by reporter identifier; see data-raw/helpers.R)
```

## Clinical end points

The character strings of the SDRF tables were processed into numeric
survival times and event indicators. For the Glas cohort this was possible
only for the set of hybridizations carrying the clinical information
(`glasRGcy5`); the putative cohort of origin of each patient (78 van 't
Veer and 84 Van de Vijver cases) was inferred from the suffix of the scan
names.

```{r genPheno, eval = FALSE}
pheno <- glasRGcy5$targets
pheno$putativeCohort <- factor(nchar(gsub(".+_", "", pheno$Scan.Name)),
  labels = c("putativeVantVeer", "putativeVanDeVijver")
)
pheno$OS <- as.numeric(gsub("\\s.+", "", pheno$Factor.Value.overall_survival))
pheno$OSevent <- as.numeric(pheno$Factor.Value.Event_Death)
pheno$TenYearSurv <- pheno$OS < 10 & pheno$OSevent == 1
TTM <- pheno$Factor.Value.Time_to_development_of_distant_metastases
pheno$TTM <- as.numeric(gsub("\\s.+", "", TTM))
pheno$TTMevent <- as.numeric(pheno$Factor.Value.Event_distant_metastases)
pheno$TTM[is.na(pheno$TTM)] <- pheno$OS[is.na(pheno$TTM)]
pheno$FiveYearMetastasis <- pheno$TTM < 5 & pheno$TTMevent == 1
```

The resulting prognosis groups, which correspond to the "good" and "bad"
prognosis groups used in the original studies, are summarized below.

```{r phenoSummary}
## Glas cohort: putative cohort of origin and 5-year distant metastasis
table(glasRGcy5$targets$putativeCohort)
table(glasRGcy5$targets$OSevent, useNA = "ifany")
table(
  FiveYearMetastasis = glasRGcy5$targets$FiveYearMetastasis,
  glasRGcy5$targets$putativeCohort, useNA = "ifany"
)
## Buyse cohort: 5-year distant recurrence by providing center
table(
  FiveYearRecurrence = buyseRGcy5$targets$FiveYearRecurrence,
  buyseRGcy5$targets$Characteristics.BioSourceProvider
)
table(
  MammaPrint = buyseRGcy5$targets$Factor.Value.MammaPrint.prediction,
  FiveYearRecurrence = buyseRGcy5$targets$FiveYearRecurrence
)
## patients excluded from the original analysis (unknown ER status)
table(buyseRGcy5$targets$toExclude)
```

# Session information

```{r sessionInfo, echo = FALSE}
sessionInfo()
```

# References
